---
title: "Homework 2"
author: Rachael Latimer, Makayla Whitney
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Part 1: Predicting a Categorical Outcome using Regularized Logistic Regression

```{r data setup}
# Load the following packages needed for modeling in this assignment
  
require(caret)
require(recipes)
require(finalfit)
require(glmnet)

# Import the tweet dataset with embeddings

tweet <- read.csv('https://raw.githubusercontent.com/uo-datasci-specialization/c4-ml-fall-2021/main/content/post/hw2/data/hw1_tweet_final.csv',header=TRUE)

# Recipe for the tweet dataset

blueprint_tweet <- recipe(x  = tweet,
                          vars  = colnames(tweet),
                          roles = c('outcome',rep('predictor',772))) %>%
  step_dummy('month',one_hot=TRUE) %>% 
  step_harmonic('day',frequency=1,cycle_size=7, role='predictor') %>%
  step_harmonic('date',frequency=1,cycle_size=31,role='predictor') %>%
  step_harmonic('hour',frequency=1,cycle_size=24,role='predictor') %>%
  step_normalize(paste0('Dim',1:768)) %>%
  step_normalize(c('day_sin_1','day_cos_1',
                   'date_sin_1','date_cos_1',
                   'hour_sin_1','hour_cos_1')) %>%
  step_num2factor(sentiment,
                  transform = function(x) x + 1,
                  levels=c('Negative','Positive'))

  
    # Notice that I explicitly specified role=predictor when using
    # step_harmonic function. This assures that the newly derived sin and cos
    # variables has a defined role.
    # You need to do this otherwise caret::train function breaks.
    # caret_train requires every variable in the recipe to have a role
    
    # You can run the following code and make sure every variable has a defined 
    # role. If you want to experiment, remove the role=predictor argument
    # in the step_harmonic function, create the recipe again, and run the following
    # you will see that the new sin and cos variables have NA in the column role
    # and this breaks the caret::train function later.
  
    # Also, in the last line, we transform the outcome variable 'sentiment' to 
    # a factor with labels. 
    # This seems necessary for fitting logistic regression via caret::train

    View(blueprint_tweet %>% prep() %>% summary)


```

##Task 1.1. 
###Split the original data into two subsets: training and test. Let the training data have the 80% of cases and the test data have the 20% of the cases.

```{r 1.1, echo=FALSE}



```

##Task 1.2. 
###Use the caret::train() function to train a model with 10-fold cross-validation for predicting the probability of sentiment being positive using logistic regression without any regularization. Evaluate and report the performance of the model on the test dataset.
```{r 1.2}


```

##Task 1.3. 
###Use the caret::train() function to train a model with 10-fold cross-validation for predicting the probability of sentiment being positive using logistic regression with ridge penalty. Try different values of ridge penalty to decide the optimal value. Use logLoss as a metric for optimization. Plot the results, and report the optimal value of ridge penalty.
```{r 1.3}



```

##Task 1.4. 
###Use the caret::train() function to train a model with 10-fold cross-validation for predicting the probability of sentiment being positive using logistic regression with lasso penalty. Try different values of lasso penalty to decide optimal value. Use logLoss as a metric for optimization. Plot the results, and report the optimal value of lasso penalty.
```{r 1.4}




```

##Task 1.5 
###Evaluate the performance of the models in 1.2, 1.3, and 1.4 on the test dataset. Calculate and report logLoss (LL), area under the reciever operating characteristic curver (AUC), overall accuracy (ACC), true positive rate (TPR), true negative rate (TNR), and precision (PRE) for three models. When calculating ACC, TPR, TNR, and PRE, assume that we use a cut-off value of 0.5 for the predicted probabilities. Summarize these numbers in a table like the following. Decide and comment on which model you would use to predict sentiment of a tweet moving forward.
```{r 1.5}



```

##Task 1.6 
###For the model you decided in 1.5, find and report the most important 10 predictors of sentiment and their coefficients. Briefly comment which variables seem to be the most important predictors.
```{r 1.6}


```

##Task 1.7. 
###Below are the two tweets I picked from my timeline. Use the model you decided in Task 1.5 to predict a probability that the sentiment being positive for these tweets. You are welcome to extract the word embeddings for these tweets by yourself (model: roberta-base, layer=12). Assume that all these tweets are posted on Saturday, May 1, 2021 at 12pm. For convenience, you can also download the dataset from the link below in case you have trouble in extracting the word embeddings.
```{r 1.7}



```

##Task 1.8. 
###Let’s do an experiment and test whether or not the model is biased against certain groups when detecting sentiment of a given text. Below you will find 10 hypothetical tweets with an identical structure. The only thing that changes from tweet to tweet is the subject.
###You are welcome to extract the word embeddings for these tweets by yourself (model: roberta-base, layer=12). Assume that all these tweets are posted on Saturday, May 1, 2021 at 12pm. For convenience, you can also download the dataset from the link below in case you have trouble in extracting the word embeddings.
###Use your model from Task 1.5 to predict the probability of these hypothetical tweets having a positive sentiment, and report these numbers in a table.
```{1.8}




```

What do you think? Does your model favor any group or seem to be biased against any group? Provide a brief commentary (not more than 500 words).



#Part 2: Predicting a Continous Outcome using Regularized Linear Regression

```{r load data}
# Load the following packages needed for modeling in this assignment
  
require(caret)
require(recipes)
require(finalfit)
require(glmnet)

# Import the oregon dataset

oregon <- read.csv('https://raw.githubusercontent.com/uo-datasci-specialization/c4-ml-fall-2021/main/content/post/hw2/data/hw1_oregon_final.csv',header=TRUE)

# Recipe for the oregon dataset

  outcome <- 'score'
  
  id      <- 'id'

  categorical <- c('sex','ethnic_cd','tst_bnch','migrant_ed_fg','ind_ed_fg',
                   'sp_ed_fg','tag_ed_fg','econ_dsvntg','stay_in_dist',
                   'stay_in_schl','dist_sped','trgt_assist_fg',
                   'ayp_dist_partic','ayp_schl_partic','ayp_dist_prfrm',
                   'ayp_schl_prfrm','rc_dist_partic','rc_schl_partic',
                   'rc_dist_prfrm','rc_schl_prfrm','grp_rpt_dist_partic',
                   'grp_rpt_schl_partic','grp_rpt_dist_prfrm',
                   'grp_rpt_schl_prfrm')

  numeric <- c('enrl_grd')

  cyclic <- c('date','month')


blueprint_oregon <- recipe(x     = oregon,
                    vars  = c(outcome,categorical,numeric,cyclic),
                    roles = c('outcome',rep('predictor',27))) %>%
  step_indicate_na(all_of(categorical),all_of(numeric)) %>%
  step_zv(all_numeric()) %>%
  step_impute_mean(all_of(numeric)) %>%
  step_impute_mode(all_of(categorical)) %>%
  step_harmonic('date',frequency=1,cycle_size=31,role='predictor') %>%
  step_harmonic('month',frequency=1,cycle_size=12,role='predictor') %>%
  step_ns('enrl_grd',deg_free=3) %>%
  step_normalize(c(paste0(numeric,'_ns_1'),paste0(numeric,'_ns_2'),paste0(numeric,'_ns_3'))) %>%
  step_normalize(c("date_sin_1","date_cos_1","month_sin_1","month_cos_1")) %>%
  step_dummy(all_of(categorical),one_hot=TRUE) %>%
  step_rm(c('date','month'))
    
  View(blueprint_oregon %>% prep() %>% summary)

```

##Task 2.1. 
###Check the dataset for missingness. If there is any variable with more than 75% missingness, remove these variables.
```{r 2.1}

missing <- ff_glimpse(oregon)$Continuous

head(missing)

#flagging missing variables 

flag_na <- which(as.numeric(missing$missing_percent) > 75)
flag_na

# Remove the flagged variables with more than 75% missingness

ordata <- oregon[,-flag_na]



```

##Task 2.2. 
###Split the original data into two subsets: training and test. Let the training data have the 80% of cases and the test data have the 20% of the cases.
```{r 2.2}

set.seed(11152021)  # for reproducibility
  
loc      <- sample(1:nrow(ordata), round(nrow(ordata) * 0.8))
ordata_train  <- ordata[loc, ]
ordata_test  <- ordata[-loc, ]

#train the blueprint
prepare <- prep(blueprint_oregon, 
                training = ordata_train)
prepare

```

##Task 2.3. 
###Use the caret::train() function to train a model with 10-fold cross-validation to predict the scores using linear regression without any regularization. Evaluate the performance of the model on both training and test datasets. Evaluate and report RMSE, R-square, and MAE for both training and test datasets. Is there any evidence of overfitting?
```{r 2.3}



```

##Task 2.4. 
###Use the caret::train() function to train a model with 10-fold cross-validation to predict the scores using ridge regression. Try different values of lambda to decide optimal value. Evaluate the performance of the model on the test dataset, and report RMSE, R-square, and MAE. Does ridge regression provide any improvement over linear regression with no regularization?
```{r 2.4}



```

##Task 2.5. 
###Use the caret::train() function to train a model with 10-fold cross-validation to predict the scores using lasso regression. Try different values of lambda to decide optimal value. Evaluate the performance of the model on the test dataset, and report RMSE, R-square, and MAE. Does lasso regression provide any improvement over linear regression with no regularization?
```{r 2.5}



```

##Task 2.6 
###Evaluate the performance of the models in 2.2, 2.3, and 2.4 on the test dataset. Calculate and report the root mean squared error (RMSE), mean absolute error (MAE), and R-square. Summarize these numbers in a table like the following. Decide and comment on which model you would use to predict scores.
```{r 2.6}




```

##Task 2.7 
###For the model you decided in 2.6, find and report the most important 10 predictors of test scores and their regression coefficients. Briefly comment which variables seem to be the most important predictors.
```{r 2.7}



```

